Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-26 tests-aniso-conduct.tex
Edited subsections 11.2.11 through 11.2.14.

In a plasma, conduction takes place primarily due to electron motion.
In astrophysical environments such as the intergalactic and
intracluster media, with typical magnetic strengths of $\sim 1$~$\mu$G
and temperatures in the millions of Kelvin, the electron gyroradius is
very small compared to the scales of interest in a typical
cosmological simulation.  As a result, electrons can only move
\textit{along} magnetic field lines, not perpendicular to them, which
can result in very interesting magnetohydrodynamical instabilities
\citep[e.g.,][]{2008ApJ...677L...9P,2008ApJ...688..905P}.  The correct
modeling of this behavior can be quite challenging, and is described
in Section~\ref{sec.num.conductions}.  Conduction in \enzo\ is
calculated in both an operator-split and directionally-split manner,
and the addition of heat transport solely along magnetic field lines
requires computation of cross-terms in the temperature derivative at
cell faces, which can add spurious oscillations in the temperature
field in regions where the temperature gradient is strong in more than
one spatial direction unless an appropriate flux-limiter \citep[such
as that of][]{1977JCoPh..23..263V} is chosen.

Figure~\ref{fig.conduct} shows a test that demonstrates the correct
behavior of anisotropic thermal conduction in \enzo.  We initialize a
two-dimensional, $256 \times 256$ cell simulation having a physical
scale of 1 kpc on a side, a uniform density of 1 proton/cc, and a
background temperature of $10^6$ Kelvin.  Magnetic fields with a
strength of B$_0 = 1$~$\mu$G are initialized such that the field lines
form circles around the center of the simulation volume, such that
B$_x = -B_0\sin(\theta)$ and B$_y = B_0\cos(\theta)$, where $\theta$
is the angle measured from the $+x$ direction in a counterclockwise
manner.  A Gaussian temperature pulse is injected at $(0.75, 0.5)$ (in
units of the box size), with a peak temperature of $10^8$ K and a FWHM
of $1/64$ of the box size.  This initial setup is shown in the left
panel of Figure~\ref{fig.conduct}.  The simulation is then allowed to
evolve with \textit{only} anisotropic conduction turned on (e.g., no
hydrodynamics, radiative cooling, or cosmological expansion), and with
a Spitzer fraction of f$_{sp} = 1$.

The right panel of Figure~\ref{fig.conduct} shows the state of the
simulation after 300 Myr.  Heat has clearly been transported only along
field lines -- there has been no diffusion perpendicular to the
magnetic field setup, which is critical for many studies involving
anisotropic thermal conduction.  No oscillations are seen in the
temperature field in regions where the fields are not aligned with the
grid, suggesting that the flux-limiter is operating as expected.

\begin{figure}
\begin{center}
\includegraphics[width=0.42\textwidth]{figures/aniso_conduction_initial_output.eps}
\includegraphics[width=0.4\textwidth]{figures/aniso_conduction_final_output.eps}
\caption{Two-dimensional anisotropic conduction test in a uniform,
constant temperature background with circular magnetic fields centered
on (0.5, 0.5) in a $256 \times 256$ grid.  The background medium has a
density of 1 proton/cc and a temperature of $10^6$~K.  At t$ = 0$
(left panel), a Gaussian heat pulse is injected at (0.75, 0.5) with a
FWHM of $1/64$ (with all numbers given in units of the box size) and a
peak temperature of $10^8$~K, and allowed to evolve without
hydrodynamical motion (i.e., static gas) and no radiative cooling for
300 Myr.  At t$ = 300$~Myr (right panel), heat has been transported
along magnetic field lines with no significant diffusion perpendicular
to field lines. Furthermore, there are no detectable oscillations in
the temperature in regions where the magnetic field is not parallel
with the grids.}
\label{fig.conduct}
\end{center}
\end{figure}

>>>>

In a plasma, conduction takes place primarily due to electron motion.
In astrophysical environments such as the intergalactic and
intracluster media, with typical magnetic strengths of $\sim 1$~$\mu$G
and temperatures in the millions of Kelvin, the electron gyroradius is
very small compared to the scales of interest in a typical
cosmological simulation.  As a result, electrons can only move
\textit{along} magnetic field lines, not perpendicular to them, which
can result in very interesting magnetohydrodynamical instabilities \protect \citep[e.g.,][]{2008ApJ...677L...9P,2008ApJ...688..905P}. The correct
modeling of this behavior can be quite challenging, and is described
in Section~\protect\ref{sec.num.conductions}. Conduction in \protect\enzo\ is calculated in both an operator-split and directionally-split manner,
and the addition of heat transport solely along magnetic field lines
requires computation of cross-terms in the temperature derivative at
cell faces, which can add spurious oscillations in the temperature
field in regions where the temperature gradient is strong in more than
one spatial direction unless an appropriate flux-limiter \protect\citep[such as that of][]{1977JCoPh..23..263V} is chosen.
\mqk{The previous paragraph more or less repeats what's in Section~\ref{sec.num.conductions}, I think it should be removed (or merged with that section).}

Figure~\ref{fig.conduct} shows a test that demonstrates the correct
behavior of anisotropic thermal conduction in \enzo.  We initialize a
two-dimensional, $256 \times 256$ cell simulation having a physical
scale of 1 kpc on a side, a uniform density of 1 proton cm$^{-3}$ and a
background temperature of $10^6$ Kelvin.  Magnetic fields with a
strength of B$_0 = 1$~$\mu$G are initialized such that the field lines
form circles around the center of the simulation volume, such that
B$_x = -B_0\sin(\theta)$ and B$_y = B_0\cos(\theta)$, where $\theta$
is the angle measured from the $+x$ direction in a counterclockwise
manner.  A Gaussian temperature pulse is injected at $(0.75, 0.5)$ (in
units of the box size), with a peak temperature of $10^8$ K and a FWHM
of $1/64$ of the box size.  This initial setup is shown in the left
panel of Figure~\ref{fig.conduct}.  The simulation is then allowed to
evolve with \textit{only} anisotropic conduction turned on (e.g., no
hydrodynamics, radiative cooling, or cosmological expansion), and with
a Spitzer fraction of f$_{\rm sp} = 1$.

The right panel of Figure~\ref{fig.conduct} shows the state of the
simulation after 300 Myr.  Heat has clearly been transported only
along field lines -- there has been no diffusion perpendicular to the
magnetic field setup, which is critical for many studies involving
anisotropic thermal conduction.  No oscillations are seen in the
temperature field in regions where the fields are not aligned with the
grid, suggesting that the flux-limiter is operating as expected (see
discussion in Section~\ref{sec.num.conductions}).

\begin{figure}
\begin{center}
\includegraphics[width=0.42\textwidth]{figures/aniso_conduction_initial_output.eps}
\includegraphics[width=0.4\textwidth]{figures/aniso_conduction_final_output.eps}
\caption{Two-dimensional anisotropic conduction test in a uniform,
constant temperature background with circular magnetic fields centered
on (0.5, 0.5) in a $256 \times 256$ grid. \mqk{Maybe show some field
lines in the figure?} The background medium has a density of 1
proton/cc and a temperature of $10^6$~K.  At t$ = 0$ (left panel), a
Gaussian heat pulse is injected at (0.75, 0.5) with a FWHM of $1/64$
(with all numbers given in units of the box size) and a peak
temperature of $10^8$~K, and allowed to evolve without hydrodynamical
motion (i.e., static gas) and no radiative cooling for 300 Myr.  At t$
= 300$~Myr (right panel), heat has been transported along magnetic
field lines with no significant diffusion perpendicular to field
lines. Furthermore, there are no detectable oscillations in the
temperature in regions where the magnetic field is not parallel with
the grids.}
\label{fig.conduct}
\end{center}
\end{figure}

### 2013-05-26 tests-onezonecollapse.tex
Edited subsections 11.2.11 through 11.2.14.

\begin{equation}
\frac{d\rho}{dt} = \frac{\rho}{t_{col}},
\end{equation}

where the collapse timescale, $t_{col}$, is

\begin{equation} \label{eqn.tcol}
t_{col} = \frac{t_{dyn}}{\sqrt{1 - f}},
\end{equation}
In this equation, $t_{dyn}$ is the dynamical time and is expressed as
\begin{equation}
t_{dyn} = \sqrt{\frac{3 \pi}{32 G \rho}}.
\end{equation}

with G being the gravitational constant.  The collapse timescale is
altered from the dynamical time by a factor $f$ in Equation
\ref{eqn.tcol}, which is an approximation of the ratio of the gas
pressure to the force of gravity.  The value of $f$ depends on the
effective adiabatic index, $\gamma_{ef} \equiv (\partial ln\ p
/ \partial ln\ \rho)$, which we calculate by linearly extrapolating
from derivative values for the two previous timesteps.  For the value
of $f$ in this test problem, we use the piecewise function of
\citet{2005ApJ...626..627O}, given by

\begin{equation}
f = \left\{
  \begin{array}{ll}
  0, & \gamma_{ef} < 0.83,\\
  0.6 + 2.5 (\gamma_{ef} - 1) - 6.0 (\gamma_{ef} - 1)^{2}, & 0.83 <
  \gamma_{ef} < 1,\\
  1.0 + 0.2 (\gamma_{ef} - 4/3) - 2.9 (\gamma_{ef} - 4/3)^{2}, & \gamma_{ef} > 1.
\end{array} \right.
\end{equation}

@@

where $\Lambda$ is the cooling rate in units of erg s$^{-1}$ g$^{-1}$
and energy, temperature, density, and pressure are related by the
ideal gas law.  Figure \ref{fig.onezone} shows an example of the
one-zone collapse test performed with an initial number density of 1
hydrogen atom per cm$^{-3}$ and an initial temperature of 100 K using
the 12 species chemistry network with H, D, and He species and metal
cooling rates calculated with the \texttt{Cloudy} code.  The effects
of metal cooling can be clearly seen; as the metallicity increases
from zero to $10^{-2}$~Z$_\odot$, the gas deviates from the primoridal
result (black line) both more rapidly and more significantly.  Our
primordial results compare very well to those shown in
\citet{2005ApJ...626..627O}; however, we use a different cooling
method than Omukai, so the lines describing the evolution of the
metal-enriched gas are not directly comparable.

\begin{figure}
  \begin{center}
    \includegraphics[width=1.0\textwidth]{figures/OneZoneCollapseTest.eps}
  \end{center}
  \caption{Evolution of temperature versus number density for a
one-zone collapse test for gas at metallicities from 0 to 10$^{-2}
Z_{\odot}$.  This test is based on the results of
\citet{2005ApJ...626..627O}, and approximates the collapse of a
self-gravitating, cooling gas cloud.  This test problem uses \enzo's
primordial chemistry network with tabulated metal cooling rates
calculated with the \texttt{Cloudy} code, and compares favorably to
Omukai's results.}
  \label{fig.onezone}
\end{figure}

>>>>

\begin{equation}
\frac{d\rho}{dt} = \frac{\rho}{t_{\rm col}},
\end{equation}

where the collapse timescale, $t_{\rm col}$, is

\begin{equation} \label{eqn.tcol}
t_{\rm col} = \frac{t_{\rm dyn}}{\sqrt{1 - f}},
\end{equation}
In this equation, $t_{\rm dyn}$ is the dynamical time and is expressed as
\begin{equation}
t_{\rm dyn} = \sqrt{\frac{3 \pi}{32 G \rho}}.
\end{equation}

with G being the gravitational constant. \mqk{Instead of explaining what $G$ is here, would it make sense to have a table at the beginning of the paper listing all the natural constants and variable symbols that are used throughout the paper?} The collapse timescale is
altered from the dynamical time by a factor $1/\sqrt{1-f}$ in Equation
\ref{eqn.tcol}, which is an approximation of the ratio of the gas
pressure to the force of gravity.  The value of $f$ depends on the
effective adiabatic index, $\gamma_{\rm ef} \equiv (\partial \ln p
/ \partial \ln \rho)$, which we linearly extrapolate
from derivative values at the two previous timesteps.  For the value
of $f$ in this test problem, we use the piecewise function of
\citet{2005ApJ...626..627O}, given by

\begin{equation}
f = \left\{
  \begin{array}{ll}
  0, & \gamma_{\rm ef} < 0.83,\\
  0.6 + 2.5 (\gamma_{\rm ef} - 1) - 6.0 (\gamma_{\rm ef} - 1)^{2}, & 0.83 <
  \gamma_{\rm ef} < 1,\\
  1.0 + 0.2 (\gamma_{\rm ef} - 4/3) - 2.9 (\gamma_{\rm ef} - 4/3)^{2}, & \gamma_{\rm ef} > 1.
\end{array} \right.
\end{equation}

@@

where $\Lambda$ is the cooling rate in units of erg s$^{-1}$ g$^{-1}$
and energy, temperature, density, and pressure are related by the
ideal gas law.  Figure \ref{fig.onezone} shows an example of the
one-zone collapse test performed with an initial number density of 1
hydrogen atom per cm$^{-3}$ and an initial temperature of 100 K using
the 12 species chemistry network with H, D, and He species and metal
cooling rates calculated with the \texttt{Cloudy} code.  The effects
of metal cooling can be clearly seen; as the metallicity increases
from zero to $10^{-2}$~Z$_\odot$, the gas rapidly and significantly
deviates from the primordial result (black line).  Our primordial
results compare very well to those shown in
\citet{2005ApJ...626..627O}; however, we use a different metal cooling
method, so the lines describing the evolution of the
metal-enriched gas are not directly comparable.

\begin{figure}
  \begin{center}
    \includegraphics[width=1.0\textwidth]{figures/OneZoneCollapseTest.eps}
  \end{center}
  \caption{Evolution of temperature versus number density for a
one-zone collapse test for gas at metallicities from 0 to 10$^{-2}
Z_{\odot}$.  This test is based on the results of
\citet{2005ApJ...626..627O}, and approximates the collapse of a
self-gravitating, cooling gas cloud.  This test problem uses \enzo's
primordial chemistry network with tabulated metal cooling rates
calculated with the \texttt{Cloudy} code, and compares favorably to
the results of \citet{2005ApJ...626..627O}.}
  \label{fig.onezone}
\end{figure}

### 2013-05-26 tests-photoevaporation.tex
Edited subsections 11.2.11 through 11.2.14.

The photo-evaporation of dense clumps of gas is prevalent in radiation
hydrodynamics simulations, and this test problem examines the
ionization front propagation into the dense clump, shadowing effects
behind the clump, and the hydrodynamic response on the clump from
photo-heating, all using \enzo's Moray ray-tracing module.  The
problem setup is the same as Test 7 in the Cosmological Radiative
Transfer Comparison Project \citep{IlievEtAl2009} and
\citet{Wise11_Moray}.  The simulation domain is 6.6~kpc on a side with
an ambient medium of pure neutral hydrogen $n_{\rm H} = 2 \times
10^{-4}\; \cubecm$ and $T = 8000 \unit{K}$.  We place a spherical
overdensity in hydrostatic equilibrium with the ambient medium.  It
has a radius $r = 0.8 \unit{kpc}$, hydrogen density $n_{\rm H} = 0.04
\cubecm$, and temperature $T = 40$ K, and it is centered at $(x,y,z) =
(5, 3.3, 3.3) \unit{kpc}$.  In \citet{IlievEtAl2009}, all of the codes
used a fixed $128^3$ grid to ease the comparison, but in this test to
demonstrate a higher resolution AMR solution, we employ a $128^3$ grid
with two additional levels of refinement with cells with a baryon mass
(method 2 in Section~\ref{sec:refinement_criteria}) greater than 1.5
being flagged for refinement.  This test is run for 15 Myr.

This cloud is subject to radiation from a point source at the center
of the $x=0$ boundary with an ionizing photon luminosity
$\dot{N}_\gamma = 3 \times 10^{51}$ photons s$^{-1}$, corresponding to
a flux $F_0 = 10^6 \unit{photons s}^{-1} \unit{cm}^{-2}$ at the clump
surface closest to the radiation source.  The radiation source has a
spectrum of a $T = 10^5 \unit{K}$ blackbody, and we use four energy
groups with the following mean energies and relative luminosities:
$E_i = (17.98, 31.15, 49.09, 76.98) \unit{eV}, L_i/L = (0.23, 0.36,
0.24, 0.06)$ that are optimized to reduce errors in the solutions with
a full spectrum and energy discretization \citep{Mirocha12}.
\citep[Note that this choice of energy groups is different
from][]{Wise11_Moray}.  We use a minimum angular resolution of 10 rays
per cell and a constant radiative transfer timestep of 25 kyr.  Figure
\ref{fig:shadowing} depicts the clump after at $t = 15$ Myr with the
outer layers expanding after being photo-heated.  It also shows the
sharp shadowing effects of the dense clump in the neutral fraction
plot that is representative of ray tracing techniques.

\begin{figure}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/code-test-shadowing.eps}
  \caption{Photo-evaporation of a dense clump using the Moray
ray-tracing module.  Clockwise from the upper left corner: slices
through the clump center of the density, temperature, neutral hydrogen
fraction, and pressure at $t=15$~Myr after the initialization of the
simulation.  A point source of radiation is at the center of the -x
boundary and illuminates the clump with a constant luminosity, casting
a clear shadow behind the clump, as seen in the neutral hydrogen
fraction.}
  \label{fig:shadowing}
\end{figure}

>>>>

The photo-evaporation of dense clumps of gas is prevalent in radiation
hydrodynamics simulations, and this test problem examines the
ionization front propagation into a dense clump, shadowing effects
behind the clump, and the hydrodynamic response on the clump from
photo-heating, all using \enzo's Moray ray-tracing module.  The
problem setup is the same as Test 7 in the Cosmological Radiative
Transfer Comparison Project \citep{IlievEtAl2009} and
\citet{Wise11_Moray}.  The simulation domain is 6.6~kpc on a side with
an ambient medium of pure neutral hydrogen of density $n_{\rm H} =
2 \times 10^{-4}\; \cubecm$ and temperature $T = 8000 \unit{K}$.  We
place a spherical overdensity in hydrostatic equilibrium with the
ambient medium.  It has a radius $r = 0.8 \unit{kpc}$, hydrogen
density $n_{\rm H} = 0.04\ \cubecm$, and temperature $T = 40$ K, and
is centered at $(x,y,z) = (5, 3.3, 3.3) \unit{kpc}$.
In \citet{IlievEtAl2009} all of the codes used a fixed $128^3$ grid to
ease the comparison, but in this test to demonstrate a higher
resolution AMR solution, we employ a $128^3$ grid with two additional
levels of refinement for cells with a baryon mass greater than 1.5
(method 2 in Section~\ref{sec:refinement_criteria}). This test is run
for 15 Myr.

This cloud is subject to radiation from a point source at the center
of the $x=0$ boundary with an ionizing photon luminosity
$\dot{N}_\gamma = 3 \times 10^{51}$ photons s$^{-1}$, corresponding to
a flux $F_0 = 10^6 \unit{photons s}^{-1} \unit{cm}^{-2}$ at the clump
surface closest to the radiation source.  The radiation source has a
spectrum of a $T = 10^5 \unit{K}$ blackbody, and we use four energy
groups with the following mean energies and relative luminosities:
$E_i = (17.98, 31.15, 49.09, 76.98) \unit{eV}, L_i/L = (0.23, 0.36,
0.24, 0.06)$ that are optimized to reduce errors in the solutions with
a full spectrum and energy discretization \citep{Mirocha12}.
(Note that this choice of energy groups is different from Wise \& Abel 2011.) \nocite{Wise11_Moray}  We use a minimum angular resolution of 10 rays
per cell and a constant radiative transfer timestep of 25 kyr.  Figure
\ref{fig:shadowing} depicts the clump at $t = 15$ Myr with the
outer layers expanding after being photo-heated.  It also shows the
sharp shadowing effects of the dense clump in the neutral fraction
plot that is representative of ray tracing techniques.

\begin{figure}
  \centering
  \includegraphics[width=1.0\textwidth]{figures/code-test-shadowing.eps}
  \caption{Photo-evaporation of a dense clump using the Moray
ray-tracing module.  Clockwise from the upper left: slices
through the clump center of density, temperature, neutral hydrogen
fraction, and pressure at $t=15$~Myr after the initialization of the
simulation.  A point source of radiation is at the center of the -x
boundary and illuminates the clump with a constant luminosity, casting
a clear shadow behind the clump, as seen in the neutral hydrogen
fraction.}
  \label{fig:shadowing}
\end{figure}
